---
title: "T2"
author: "Abraham Majluf, Izzy Tice, Rubén Figueroa"
format: html
editor: visual
---

```{r, echo = FALSE, include = FALSE}
library(here)
library(tidyverse)
set.seed(12345)

# load population & employment dbs
pop_db <- read_csv(here("model",
                        "inputs",
                        "zonal",
                        "ma_population_run97-176_2019_v20240109.csv"))

emp_db <- read_csv(here("model",
                        "inputs",
                        "zonal",
                        "ma_employment_run97-176_2019_v20240109.csv"))

# identify low income population
mod_pop_db <- pop_db |>
  filter(is_worker==1) |>
  mutate(low_inc = case_when(wage_inc < 20000 ~ 1,
                              TRUE ~ 0))

# undocumented is the whole construction sector
undoc <- sum(emp_db$`1_constr`)

# undocumented as share of low income population
undoc_in_low_inc <- undoc/sum(mod_pop_db$low_inc)

# random undocumented status assignment 
mod_pop_db <- mod_pop_db |>
  filter(low_inc == 1) |>
  mutate(undoc =  rbinom(n(), size = 1, prob = undoc_in_low_inc)) |>
  mutate(unique_id = paste0(hid, person_num)) |>
  select(unique_id, undoc)

# join database
pop_db <- pop_db |>
  mutate(unique_id = paste0(hid, person_num)) |>
  left_join(mod_pop_db, by = "unique_id") |>
  mutate(undoc = replace_na(undoc, 0))

# exclude undoc in final db
pop_db_alternative <- pop_db |>
  filter(undoc == 0) |>
  select(-unique_id, -undoc)

# new household count per block
household_per_block <- pop_db_alternative |>
  group_by(block_id) |>
  summarise(total_households_new = n_distinct(hid)) 

# join database and eliminate undoc employment in construction
emp_db_alternative <- emp_db |>
  select(-total_households) |>
  left_join(household_per_block, by = "block_id") |>
  mutate(total_households = replace_na(total_households_new,0)) |>
  select(-total_households_new) |>
  mutate(`1_constr`=0) |>
  mutate(total_jobs = rowSums(across(2:11)))

# write csvs
write_csv(emp_db_alternative, here("model", "inputs", "zonal", "ma_employment_alternative.csv"))

write_csv(pop_db_alternative, here("model", "inputs", "zonal", "ma_population_alternative.csv"))
```

## Trip Generation Comparison

```{r, echo = FALSE, include = FALSE}
# Libraries for analysis:
library(tidyverse)
library(sf)
library(DBI)
library(here)
library(leaflet)
library(maptiles)
library(tidyterra)
library(knitr)
library(tigris)

sf::sf_use_s2(FALSE)
```

### Trip Production:

First, we load the output from TransCAD for both the base and alternative scenarios:

```{r, echo = FALSE}
# Transcad output:
## Base scenario:
base_db <- dbConnect(drv=RSQLite::SQLite(), 
                     dbname=here("model",
                                 "outputs",
                                 "Base",
                                 "tdm23.db"))

base_trip_prod <- dbGetQuery(base_db, 'SELECT * FROM trip_prod')

dbDisconnect(base_db)

## Alternative scenario:
alt_db <- dbConnect(drv=RSQLite::SQLite(), 
                     dbname=here("model",
                                 "outputs",
                                 "Alternative",
                                 "tdm23.db"))

alt_trip_prod <- dbGetQuery(alt_db, 'SELECT * FROM trip_prod')

dbDisconnect(alt_db)
```

Now we analyze the differences between scenarios for trip production:

```{r}
# Agregate trip productions on the block level:
## Base scenario:
base_block_prod <- base_trip_prod |>
  replace_na(list(hbw_p = 0,
                  hbsc_p = 0,
                  hbsr_p = 0,
                  hbpb_p = 0,
                  nhbw_p = 0,
                  nhbnw_p = 0)) |>
  group_by(block_id) |>
  summarise(hbw_p = sum(hbw_p),
            hbsc_p = sum(hbsc_p),
            hbsr_p = sum(hbsr_p),
            hbpb_p = sum(hbpb_p),
            nhbw_p = sum(nhbw_p),
            nhbnw_p = sum(nhbnw_p))

# Alternative scenario:
alt_block_prod <- alt_trip_prod |>
  replace_na(list(hbw_p = 0,
                  hbsc_p = 0,
                  hbsr_p = 0,
                  hbpb_p = 0,
                  nhbw_p = 0,
                  nhbnw_p = 0)) |>
  group_by(block_id) |>
  summarise(hbw_p_alt = sum(hbw_p),
            hbsc_p_alt = sum(hbsc_p),
            hbsr_p_alt = sum(hbsr_p),
            hbpb_p_alt = sum(hbpb_p),
            nhbw_p_alt = sum(nhbw_p),
            nhbnw_p_alt = sum(nhbnw_p))

# Define the tolerance for difference identification:
tolerance <- 0.75

# Compute differences in trip generation:
compare_prod <- full_join(base_block_prod, alt_block_prod) |>
  mutate(dif_hbw_p = hbw_p_alt - hbw_p,
         dif_hbsc_p = hbsc_p_alt - hbsc_p,
         dif_hbsr_p = hbsr_p_alt - hbsr_p,
         dif_hbpb_p = hbpb_p_alt - hbpb_p) |>
  filter(abs(dif_hbw_p)/hbw_p > tolerance |
         abs(dif_hbsc_p)/hbsc_p > tolerance |
         abs(dif_hbsr_p)/hbsr_p > tolerance |
         abs(dif_hbpb_p)/hbpb_p > tolerance)

# Graphical analysis:
ma_blocks <- blocks(state = "MA", 
                          year = 2010,
                          progress_bar = FALSE) |>
  st_transform("WGS84") |>
  rename(block_id = GEOID10) |>
  select(block_id)

TAZs <- here("model",
             "inputs",
             "zonal",
             "shp",
             "CTPS_TDM23_TAZ_2017g_v202303.shp") |>
  st_read(quiet = TRUE) |>
  st_transform("WGS84") |>
  mutate(block_id = as.character(taz_id)) |>
  select(block_id)

zones <- rbind(ma_blocks, TAZs)

dif_blocks_prod <- zones |>
  right_join(compare_prod) 

hb_prod_labels <- paste0(formatC(dif_blocks_prod$dif_hbw_p, format = "f", digits = 2),
                        " change in HBW trips<br/>",
                        formatC(dif_blocks_prod$dif_hbsc_p, format = "f", digits = 2), 
                        " change in HB-school trips<br/>",
                        formatC(dif_blocks_prod$dif_hbsr_p, format = "f", digits = 2),
                        " change in HB-soc/rec trips<br/>",
                        formatC(dif_blocks_prod$dif_hbpb_p, format = "f", digits = 2),
                         " change in HB-per-bus") |>
  lapply(htmltools::HTML)

leaflet(dif_blocks_prod) |>
  addProviderTiles(provider = "CartoDB.Positron") |>
  addPolygons(weight = 2,
              color = "darkred",
              fillColor = "darkred",
              fillOpacity = 0.1,
              highlightOptions = highlightOptions(weight = 3,
                                                  fillOpacity = 0.5),
              label = hb_prod_labels)
```

### Trip Attraction:

First, we load the output from TransCAD for both the base and alternative scenarios:

```{r, echo = FALSE}
# Transcad output:
## Base scenario:
base_db <- dbConnect(drv=RSQLite::SQLite(), 
                     dbname=here("model",
                                 "outputs",
                                 "Base",
                                 "tdm23.db"))

base_trip_attr <- dbGetQuery(base_db, 'SELECT * FROM trip_attr')

dbDisconnect(base_db)

## Alternative scenario:
alt_db <- dbConnect(drv=RSQLite::SQLite(), 
                     dbname=here("model",
                                 "outputs",
                                 "Alternative",
                                 "tdm23.db"))

alt_trip_attr <- dbGetQuery(alt_db, 'SELECT * FROM trip_attr') |>
  rename_with(~ paste0("alt_", .x),
              ends_with("_a")) 

dbDisconnect(alt_db)
```

Now we analyze the differences between scenarios for trip attraction:

```{r}
# Define the tolerance for difference identification:
tolerance <- 0.75

# Compute differences in trip generation:
compare_attr <- full_join(base_trip_attr, alt_trip_attr) |>
  mutate(diff_hbw_inc1_a = alt_hbw_inc1_a - hbw_inc1_a,
         diff_hbw_inc2_a = alt_hbw_inc2_a - hbw_inc2_a,
         diff_hbw_inc3_a = alt_hbw_inc3_a - hbw_inc3_a,
         diff_hbw_inc4_a = alt_hbw_inc4_a - hbw_inc4_a,
         diff_hbsr_a = alt_hbsr_a - hbsr_a,
         diff_hbsc_a = alt_hbsc_a - hbsc_a,
         diff_hbpb_a = alt_hbpb_a - hbpb_a,
         diff_nhbw_a = alt_nhbw_a - nhbw_a,
         diff_nhbnw_a = alt_nhbnw_a - nhbnw_a) |>
  filter(abs(diff_hbw_inc1_a)/hbw_inc1_a > tolerance |
           abs(diff_hbw_inc2_a)/hbw_inc2_a > tolerance |
           abs(diff_hbw_inc3_a)/hbw_inc3_a > tolerance |
           abs(diff_hbw_inc4_a)/hbw_inc4_a > tolerance |
           abs(diff_hbsc_a)/hbsc_a > tolerance |
           abs(diff_hbsr_a)/hbsr_a > tolerance |
           abs(diff_hbpb_a)/hbpb_a > tolerance |
           abs(diff_nhbw_a)/nhbw_a > tolerance |
           abs(diff_nhbnw_a)/nhbnw_a > tolerance)

# Graphical analysis:
dif_blocks_attr <- zones |>
  right_join(compare_attr) 

attr_labels <- paste0(formatC(dif_blocks_attr$diff_hbw_inc1_a, format = "f", digits = 2),
                        " change in HBW trips (< $35k)<br/>",
                      formatC(dif_blocks_attr$diff_hbw_inc2_a, format = "f", digits = 2),
                        " change in HBW trips ($35k - $65k)<br/>",
                      formatC(dif_blocks_attr$diff_hbw_inc3_a, format = "f", digits = 2),
                        " change in HBW trips ($65k - $100k)<br/>",
                      formatC(dif_blocks_attr$diff_hbw_inc4_a, format = "f", digits = 2),
                        " change in HBW trips (> $100k)<br/>",
                        formatC(dif_blocks_attr$diff_hbsc_a, format = "f", digits = 2), 
                        " change in HB-school trips<br/>",
                        formatC(dif_blocks_attr$diff_hbsr_a, format = "f", digits = 2),
                        " change in HB-soc/rec trips<br/>",
                        formatC(dif_blocks_attr$diff_hbpb_a, format = "f", digits = 2),
                         " change in HB-per-bus<br/>",
                      formatC(dif_blocks_attr$diff_nhbw_a, format = "f", digits = 2),
                        " non-home-based work trips<br/>",
                      formatC(dif_blocks_attr$diff_nhbnw_a, format = "f", digits = 2),
                        " change in non-home-based non-work trips") |>
  lapply(htmltools::HTML)

leaflet(dif_blocks_attr) |>
  addProviderTiles(provider = "CartoDB.Positron") |>
  addPolygons(weight = 2,
              color = "darkred",
              fillColor = "darkred",
              fillOpacity = 0.1,
              highlightOptions = highlightOptions(weight = 3,
                                                  fillOpacity = 0.5),
              label = attr_labels)
```

### Non-home-based trip productions

First, we load the output from TransCAD for both the base and alternative scenarios:

```{r, echo = FALSE}
# Transcad output:
## Base scenario:
base_db <- dbConnect(drv=RSQLite::SQLite(), 
                     dbname=here("model",
                                 "outputs",
                                 "Base",
                                 "tdm23.db"))

base_trip_nhb <- dbGetQuery(base_db, 'SELECT * FROM prod_nhb')

dbDisconnect(base_db)

## Alternative scenario:
alt_db <- dbConnect(drv=RSQLite::SQLite(), 
                     dbname=here("model",
                                 "outputs",
                                 "Alternative",
                                 "tdm23.db"))

alt_trip_nhb <- dbGetQuery(alt_db, 'SELECT * FROM prod_nhb') |>
  rename_with(~ paste0("alt_", .x),
              ends_with("_p"))  

dbDisconnect(alt_db)
```

Now we analyze the differences between scenarios for non-home based trip production:

```{r}
# Define the tolerance for difference identification:
tolerance <- 0.75

# Compute differences in trip generation:
compare_nhb <- full_join(base_trip_nhb, alt_trip_nhb) |>
  mutate(diff_nhbw_p = alt_nhbw_p - nhbw_p,
         diff_nhbnw_p = alt_nhbnw_p - nhbnw_p) |>
  filter(abs(diff_nhbw_p)/nhbw_p > tolerance |
           abs(diff_nhbnw_p)/nhbnw_p > tolerance)

# Graphical analysis:
dif_blocks_nhb <- zones |>
  right_join(compare_nhb) 

nhb_labels <- paste0(formatC(dif_blocks_nhb$diff_nhbw_p, format = "f", digits = 2),
                        " non-home-based work trips<br/>",
                      formatC(dif_blocks_nhb$diff_nhbnw_p, format = "f", digits = 2),
                        " change in non-home-based non-work trips") |>
  lapply(htmltools::HTML)

leaflet(dif_blocks_nhb) |>
  addProviderTiles(provider = "CartoDB.Positron") |>
  addPolygons(weight = 2,
              color = "darkred",
              fillColor = "darkred",
              fillOpacity = 0.1,
              highlightOptions = highlightOptions(weight = 3,
                                                  fillOpacity = 0.5),
              label = nhb_labels)
```

End of project.
